 

### 2023-06-21 markdown/Thesis.Rmd
first draft finished

(the spacing is set to 1.5)

no more than 250 words for the abstract

- a description of the research question/knowledge gap – what we know and what we don’t know
- how your research has attempted to fill this gap
- a brief description of the methods
- brief results
- key conclusions that put the research into a larger context

>>>>

With the growing prominence of single-cell techniques across various omics fields, there is a pressing need to develop a standardized pipeline for proteomics data in the realm of systems biology. Unlike DNA sequencing and RNA sequencing, proteomics analysis via mass spectrometry incurs high costs in terms of labor and equipment. Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed. However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility of post-processing using the R programming language. 
Guided by publications such as SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of the proteomics data analysis pipeline was thoroughly tested and validated using both bulk and single-cell mass spectrometry datasets. The pipeline implementation leverages the power of an interactive environment built in R Shiny, empowering users to discover the gold standard for their specific dataset. 
Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction. 
The developed application served as a valuable tool in guiding the identification of significantly differentially expressed proteins in HIV cells. Moreover, it facilitated the mapping of these proteins to relevant pathways and provided additional biological contexts for a comprehensive understanding of their functional implications.
The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics. Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.

### 2023-06-21 markdown/conclusion.Rmd
first draft finished

Through this project, we have demonstrated that computational analysis can be made accessible to a wide range of biologists by providing an interactive environment. It is important to recognize that there cannot be a one-size-fits-all approach for proteomics datasets. Therefore, our application offers easy-to-interpret visualizations that allow users to assess the patterns and trends in their observed datasets. The default options provided in the application are based on successful analysis pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point for customization at multiple levels.

The developed pipeline and user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers. However, our future perspective involves deploying the application on a server, making it accessible to research groups around the world. This would enhance collaboration and facilitate the sharing of insights and findings across different geographical locations. By providing a centralized platform for proteomics analysis, we aim to foster scientific advancements and enable researchers to make significant contributions to their respective fields. Furthermore by having a central hub for proteomic data analysis and the use of artificial intelligence the biological meaning could be brought to a wider context. Since proteomic data could be saved without violating data protection laws, algorithms could be employed to learn from large datasets.

>>>>

Through this project, I demonstrated that computational analysis can be made accessible to a wide range of biologists by providing an interactive environment. It is important to recognize that there cannot be a one-size-fits-all approach for proteomics datasets. Therefore, the application offers easy-to-interpret visualizations that allow users to assess the patterns and trends in their observed datasets. The default options provided in the application are based on successful analysis pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point for customization at multiple levels.

The example dataset provided showcases the inherent challenges in proteomic data analysis. It exhibits skewed distributions, batch effects, and a substantial number of missing values. Despite these complexities, the developed pipeline and user interface offer biologists a convenient means to address their research questions and validate the computations performed. The example use case also sheds light on the fact that previous proteomics publications, such as the ScoPE2 study \citep{Specht2021}, may not always employ the gold standard in data processing. This emphasizes the importance of critically evaluating and refining data processing steps through trial and error to achieve optimal results. The interactive environment provided by the application allows users to navigate the trial and error process without experiencing frustration. This user-friendly interface ensures that researchers can iteratively refine their data processing steps and analyze the results in a seamless manner. Furthermore, the scalability of the application was initially unknown until the final submission of the single-cell dataset. The programming and testing of the application primarily focused on bulk datasets, typically consisting of up to four runs at once. However, it is noteworthy that the application demonstrates remarkable efficiency and performance, even when running large datasets. The ability of the application to efficiently handle larger datasets, such as single-cell studies, is a significant advantage. It allows researchers to analyze complex and extensive datasets without sacrificing performance or compromising the quality of the results. This scalability makes the application a valuable tool for analyzing diverse types of biological data, accommodating the needs of researchers working with both smaller and larger datasets.

The developed pipeline and user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers. However, one of the future perspectives involves deploying the application on a server, making it accessible to research groups around the world. This would enhance collaboration and facilitate the sharing of insights and findings across different geographical locations. By providing a centralized platform for proteomics analysis, aiming to foster scientific advancements and enable researchers to make significant contributions to their respective fields. Furthermore by having a central hub for proteomic data analysis and the use of artificial intelligence the biological meaning could be brought to a wider context. Since proteomic data could be saved without violating data protection laws, algorithms could be employed to learn from large datasets.

An important aspect of learning from the data involves annotating specimens, particularly in the context of single-cell proteomics. The clustering approach is a primary focus for future development and analysis in the near term. By employing clustering algorithms and stratifying the data, individual cells can be labeled and annotated based on their phenotypes. This annotation process involves associating protein sets collections with specific clusters or cell types, providing a more comprehensive understanding of cellular heterogeneity and functional characteristics.

In conclusion, the developed pipeline and user interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations. Future perspectives include deploying the application on a server for global accessibility, leveraging artificial intelligence techniques, and implementing clustering approaches to further enhance annotation and understanding of cellular heterogeneity and functional characteristics.
